\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-17 paper.tex
Thanks Graham
Closes #451
Pass through "diversity" section

\section{A diversity of structures}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring}
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
One source of differences between 
the inferences is the underlying model. For instance, 
the ARG in Fig.~\ref{fig-inferred-args}A is inferred 
under parsimony, and therefore has many fewer 
ancestral nodes then Fig.~\ref{fig-inferred-args}B, which 
is sampled from the coalescent with recombination with an 
appropriate recombination rate parameter. Such differences
are well understood.

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}

[WIP] The differences between ARGs that we wish to draw attention to here 
are more subtle

As is well-known (cite, perhaps Gusfield?), many different ancestral graph
topologies are compatible with the same resulting sequence data. Hence it is
unsurprising that the four graphs differ, for example in the numbers of nodes
and their relationship to each other. However, more detailed inspection of
Fig.~\ref{fig-inferred-args} reveals a diversity of representation that is only
possible via the gARG encoding. For instance, although the most parsimonious
ARG in Fig.~\ref{fig-inferred-args}A has far fewer ancestral events than an ARG
instance sampled from from the coalescent with recombination in
Fig.~\ref{fig-inferred-args}B, both consist only of nodes with either two
parents or two children (but not both). These nodes correspond to recombination
and common ancestor events in the equivalent eARG. In contrast, more scalable
heuristic inference methods (Fig.~\ref{fig-inferred-args}C and D) create ARGs
in which a node can simultaneously have more that than two children and more
than two parents. Note also that the first three methods infer nodes that have
a ``coalescence span'' of less than 100\%, indicating that in the local trees,
ancestral lineages exist that pass through these nodes without the nodes
marking a coalescent point. Such ``locally-unary'' nodes are an notable feature
of the local trees defined by an ARG (see Appendix
\ref{sec-locally-unary-edges}).

These differences, in particular the number of parents and children, and the
coalescence span, reflect the degree of precision represented in the inferred
ARGs. Indeed, it is important to note that the gARG encoding allows a single
underlying genetic genealogy to be represented at differing levels of detail.
The concept is illustrated in Fig.~\ref{fig-simplification}, which shows
different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification})
of the same underlying simulated ARG. The full ARG, with all coalescent and
recombination events represented by separate genomes, is shown in
Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
multiple recombinations into a single child or multiple coalescences into a
single parent (Fig.~\ref{fig-simplification}C, also see
Appendix~\ref{sec-cell-lineages-and-args}). Finally, if we chose to represent
only local coalescence points (Fig.~\ref{fig-simplification}D), we end up with
a ``fully simplified'' ARG of the sort inferred in
Fig.~\ref{fig-inferred-args}D. Note that in all cases, the equivalent trees
only differ in the locally-unary nodes they contain; topological relationships
between the samples remain identical.

@@

%%% Pasted from previous para

The ability to represent an ARG to differing degrees of precision is a powerful
feature of gARGs. In particular, when inferring ARGs from genome
sequencing data, the timing, positions, and even the number of recombination
events is generally not possible to infer precisely. For example, under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination
rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. Because a gARG
can encode a diversity of ARG structures, it
allows this fundamental uncertainty in inference to be appropriately represented
(Appendix~\ref{sec-precision}.
% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
%The fact that the eARG encoding \emph{requires}
%precise information about recombination is therefore a fundamental limitation.

@@

Tremendous progress has been made in recent years on the long-standing
problem of ARG inference, there is now a range of practically 
applicable methods available.
Methods targetting large-scale datasets tend to simplify
the inference problem by
making a single, deterministic 
best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast,
and it is therefore not clear that generating many guesses
when sample sizes are large will achieve much in terms of capturing uncertainty.
An alternative approach to 
is to incorporate uncertainty encountered during inference into the returned ARG.
The gARG encoding described here enables particular kinds of uncertainty
to be incorporated directly into the topology:
nodes that have more than two children (polytomies)
represent uncertainty over the ordering of coalescence events 
(Appendix~\ref{sec-cell-lineages-and-args}), 
and those have have more than two parents 
represent uncertainty over the ordering of multiple recombination events
(Appendix~\ref{sec-ARG-simplification}). 
Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
% How this uncertainty can be utilised in downstream applications
% is an open question.

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughn
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.

>>>>

\section{A diversity of structures}
\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate, converted 
to the standard \emph{tskit} gARG encoding. See
Appendix \ref{sec-example-inferred-args} for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}
A key goal of this perspective is to highlight the heterogeneity of
the graph structures inferred by modern ARG inference methods.
To illustrate this point, Fig.~\ref{fig-inferred-args} shows the 
output of 
\kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome},
\tsinfer~\citep{kelleher2019inferring}
and \relate~\citep{speidel2019method}
on the classical \cite{kreitman1983nucleotide} dataset. 
The ARGs in Fig.~\ref{fig-inferred-args}A and B are precise 
estimates (Appendix~\ref{sec-precision}),
with each node corresponding to a common ancestor 
or recombination event, or  equivalently, either having two children
or two parents.
In contrast the ARGs in Fig.~\ref{fig-inferred-args}C and D do not 
have this clear-cut interpretation, and the nodes
can simultaneously have more that than two children and more
than two parents. Another dimension of variability among the ARGs
is that the first three methods infer nodes that have
a ``coalescence span''  greater than 0 and less than 100\%, indicating that there 
are nodes that are ``locally unary'' (Appendix~\ref{sec-locally-unary-edges}),
but mark a coalescence between lineages elsewhere along the sequence.

@@

A key feature of the gARG encoding is that it enables these varying levels 
of precision to be represented, and brings these nuanced features to light.
These ideas are illustrated in Fig.~\ref{fig-simplification}, which shows
different levels of ``simplification'' (Appendix~\ref{sec-ARG-simplification})
of the same underlying simulated ARG. The full ARG, with all coalescent and
recombination events represented by separate genomes, is shown in
Fig.~\ref{fig-simplification}A. Simpler representations can be formed by
removing ``unknowable'' nodes (Fig.~\ref{fig-simplification}B) and collapsing
multiple recombinations into a single child or multiple coalescences into a
single parent (Fig.~\ref{fig-simplification}C). 
Finally, Fig.~\ref{fig-simplification}D is a ``fully simplified''
ARG, in which only coalescences in local trees are retained.
Note that while ARGs of this type (produced by default by the \texttt{msprime}
simulator, for example) lack a significant level of detail, they still
retain the key feature of shared node identity across local trees.

This ability to represent an ARG to differing degrees of precision is a powerful
feature. In particular, when inferring ARGs from genome
sequencing data, the timing, positions, and even the number of recombination
events is generally not possible to infer precisely. For example, under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation and recombination
rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. Because a gARG
can encode a diversity of ARG structures, it
allows this fundamental uncertainty in inference to be appropriately represented
(Appendix~\ref{sec-precision}).

% For general
% ARGs, for instance those representing pedigrees, or genealogies of organisms with
% low mutation rates and complex patterns of recombination, such inference can be
% even more challenging.
%The fact that the eARG encoding \emph{requires}
%precise information about recombination is therefore a fundamental limitation.

@@

Tremendous progress has been made in recent years on the long-standing
problem of ARG inference, there is now a range of practically 
applicable methods available.
Methods targetting large-scale datasets tend to simplify
the inference problem by
making a single, deterministic 
best-guess~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast,
and it is therefore not clear that generating many guesses
when sample sizes are large will achieve much in terms of capturing uncertainty.
An alternative approach to 
is to incorporate uncertainty encountered during inference into the returned ARG.
The gARG encoding described here enables particular kinds of uncertainty
to be incorporated directly into the topology:
nodes that have more than two children (polytomies)
represent uncertainty over the ordering of coalescence events 
(Appendix~\ref{sec-cell-lineages-and-args}), 
and those have have more than two parents 
represent uncertainty over the ordering of multiple recombination events
(Appendix~\ref{sec-ARG-simplification}). 
Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be utilised in downstream applications
is an open question.

@@

\section*{Acknowledgements}
We are grateful to Nick Barton, Gideon Bradburd, Alex Lewanski,
Peter Ralph and Andrew Vaughn
for helpful discussions and comments on the manuscript, and
to Iain Mathieson and Aylwyn Scally for interesting discussions that
originally motivated this work.
We are particularly grateful to Graham Coop for detailed feedback
on and earlier version of the manuscript.
% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
Jerome Kelleher acknowledges support from the Robertson Foundation,
NIH (research grants HG011395 and HG012473) and
EPSRC (research grant EP/X024881/1).
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
